Thread Management for High Performance Database Systems - Design and Implementation \\{\large Technical Report

### 2018-08-17 Presentations/Paper/paper_thread_pool.tex
Grammar and typos eliminated

\begin{abstract}
        Processing tasks in parallel is used in nearly all applications to keep
        up with the requirements of modern software systems. However, the
        current implementation of parallel processing in GeckoDB, a graph database system developed in our group, requires
        to spawn many short-lived threads that execute a single task and then
        terminate. This creates a lot of overhead since the threads are not
        being reused. To counter this effect we implemented a thread pool to be
        used in GeckoDB to reduce the additional time required
        to create and close thousands of threads. In this paper, we show our
        implementation of a thread pool to process independent tasks in parallel,
        including the waiting for a set of processed tasks. We also compare the
        thread pool implementation against the current one. Additionally, we show that the task and thread pool configuration,
        depending on the use case, has a high impact on the thread
        pool performance. At the end of our evaluation, we show that the implementation 
        fulfils all given requirements and generally reduce
        overhead compared to creating and spawning threads individually.
\end{abstract}

@@

The growing number of cores for both traditional CPUs as well as trending co-processors like GPUs require to re-evaluate core design descicion in data-intense systems. For instance, Wu et al. examined the scalability of multi-version concurrency control (MVCC) for a growning number of threads in  2017\cite{wu2017empirical}. They concluded that multi-threading in particular and modern hardware in general both promise notable performance gains for high-performance database systems using MVCC. However, they also observed that there is no clear winner combination for several concurrency control protocol, version storage, and garbage collection as well as the index management strategies in MVCC.

For this paper, we focus on managing a far more lower level: general parallel data processing. In particular, with this paper we provide insights into our thread managing strategy in GeckoDB/BOOLSTER, its design and its implementation.

@@

Each task can be enqueued with a priority. The priority zero is the highest and the task will be processed by the next free thread. The higher the priority of a task is, the further it is placed behind in the queue. Additionally, we only consider not preemtable tasks. Once a task is assigned to a thread, the thread will finish the operation of the task before getting a new one.

@@

Thread safe access to memory can be achieved by different approaches. Firstly, a mutex thread can manage the access for multiple other threads. Secondly, atomic operations ensure the safe access from different threads to the same memory. We use both approaches for different parts of the thread pool. Since using a mutex thread can decrease access performance, we decided to execute atomic operations on memory containing the thread state informations. We use a mutex thread to synchronize the enqueueing of tasks into the priority queue.

@@

Since $openTaskCount$ has to be altered from different threads the required decrements are ensured to be atomic. This is in opposition to the usual goal of slot-maps to improve cache locality. Atomics on x86 usually prevent their whole cache line from being used while they are changed. Thus a sparse distribution of slots should reduces collisions. To quantify this difference the baseline implementation searches for free slots linearly from the beginning, while an alternative approach tries to distribute the slots over multiple cache lines. In Figure \ref{fig10}, we show implementations for both search approaches.

@@

\subsection{Managing Thread States}
As mentioned before, accessing the same memory location from different threads may lead to race conditions, which means the program behaviour depends on the accessing order of the threads. For example, while one thread updates data at a specific memory location, another thread writes into the data, which results to inconsistency. In Section \ref{sec:pre}, we presented two approaches to handle multithreaded access to the same memory location, mutex threads and atomic operations and we state where those approaches are used in our implementation. In this subsection, we focus on atomic operations to manage thread states. We consider the following steps for a thread to update its state. At first, the current state has to be checked and, depending on the current state, it has to be updated. If these steps are implemented using an if statement and an update function, the complete procedure is not atomic. Consequently, another thread can update the state between check of the current state and update. To avoid this, we use the function \emph{atomic\_compare\_exchange\_strong} \cite{atomicstrong} of the stdatomic library. This function combines the check and update in one atomic procedure. We implemented every thread state access with this function to ensure atomic thread state updates.

@@

\begin{table}[htbp]
	\caption{Thread statistics with their internal names and descriptions}
	\begin{center}
		\begin{tabular}{ l l }
			\hline
			\textbf{Statistic name}&\textbf{Description}\\
			\hline
			waiting\_time & Time the thread spend waiting (ms) \\
			busy\_time & Time the thread spend executing tasks (ms)\\
			task\_count & Number of tasks the thread has executed \\
			\hline
		\end{tabular}
		\label{tab3}
	\end{center}
\end{table}

@@

For testing purposes we used primarly Linux setups, for one Ubuntu 17.04
(linux 4.13) in an virtual machine environment, second Archlinux (linux
4.17.3-1) running native, both updated on stable branch. Tests have also been
conducted on Windows 10 and macOs HighSierra machines.

In our setups Sky- and Kabylake aswell as Ryzen processors have been used with memory
ranging from 8 to 24 GB of both DDR3-SODIMM and DDR4-SODIMM. 
All benchmarks have been repeated three times and a median value of all
repetitions is being used for evaluation.

@@

\subsection{Validity of Results} 
Before presenting our evaluation results, we
give some preliminaries about the validity of our results. At first, all
measurements are executed on all machines we listed in our experimental
configuration, but in this work we present only the results measured on
Archlinux with kernel 4.17.3-1 and 24 GB DDR4-SODIMM. We decided to use this
system, since it is running a current kernel version and not in a virtual
environment. Additionally, we use gnuplot to generate diagrams from our results,
consequently using a windows machine is not optimal.

@@

\subsection{Waiting Time Evaluation}
As mentioned before, instead of implementing explicit scheduling to assign tasks to threads, we use the \emph{first come first serve} approach. Each thread that currently has no task, checks the priority queue to pop a new task in a loop. Since our priority queue manages the memory access from different threads using a mutex thread, the queue is locked while a thread asks for a task. Consequently, in large thread pools, the accessing time of the priority queue may increase along with the waiting time of threads. To examine this behaviour with increasing thread pool size, we measure the average waiting time of thread pools with different sizes.

@@

\subsubsection{Benchmark Implementation}
To examine the waiting time behaviour of larger thread pools, we use a similar implementation than the \emph{cmp\_pool} method, which creates 10000 tasks at different thread pool sizes. Again, we consider sizes from $4$ to $32$ threads as useful test case and take the average value of three measurements. To measure waiting and completion time, we use the monitoring component of our system and evaluate the thread pool statistics of each instance.

@@

\section{Related Work}\label{sec:related}
Implementing multi threading can be realised using different approaches from spawning new threads to reuse them in thread pools. Pyarali et al. show that a thread pool model can significantly improve system performance and reduce response time \cite{pyarali2001evaluating}. They present different approaches to implement thread pools and show the dependency on use cases of these implementations.

@@

\section{Conclusion and Future Work}\label{sec:future}
In this work, we describe our implementation of a thread pool system. We present in detail the implementation and show our evaluation results containing a comparison against the baseline implementation and the waiting time behaviour over different thread pool sizes. Within our implementation, we state the differences of parallel memory access as well as the problem of implementing synchronous waiting for a set of tasks.

>>>>

\begin{abstract}
        Processing tasks in parallel is used in nearly all applications to keep
        up with the requirements of modern software systems. However, the
        current implementation of parallel processing in GeckoDB, a graph database system developed in our group, requires
        to spawn many short-lived threads that execute a single task and then
        terminate. This creates a lot of overhead since the threads are not
        being reused. To counter this effect, we implemented a thread pool to be
        used in GeckoDB to reduce the additional time required
        to create and close thousands of threads. In this paper, we show our
        implementation of a thread pool to process independent tasks in parallel,
        including the waiting for a set of processed tasks. We also compare the
        thread pool implementation against the current one. Additionally, we show that the task and thread pool configuration,
        depending on the use case, has a high impact on the thread
        pool performance. At the end of our evaluation, we show that the implementation 
        fulfils all given requirements and generally reduce
        overhead compared to creating and spawning threads individually.
\end{abstract}

@@

The growing number of cores for both traditional CPUs as well as trending co-processors like GPUs require to re-evaluate core design decision in data-intense systems. For instance, Wu et al. examined the scalability of multi-version concurrency control (MVCC) for a growing number of threads in  2017\cite{wu2017empirical}. They concluded that multi-threading in particular and modern hardware in general both promise notable performance gains for high-performance database systems using MVCC. However, they also observed that there is no clear winner combination for several concurrency control protocol, version storage, and garbage collection as well as the index management strategies in MVCC.

For this paper, we focus on managing a far lower level: general parallel data processing. In particular, with this paper we provide insights into our thread managing strategy in GeckoDB/BOOLSTER, its design and its implementation.

@@

Each task can be enqueued with a priority. The priority zero is the highest and the task will be processed by the next free thread. The higher the priority of a task is, the further it is placed behind in the queue. Additionally, we only consider non-preemptable tasks. Once a task is assigned to a thread, the thread will finish the operation of the task before getting a new one.

@@

Thread safe access to memory can be achieved by different approaches. Firstly, a mutex thread can manage the access for multiple other threads. Secondly, atomic operations ensure the safe access from different threads to the same memory. We use both approaches for different parts of the thread pool. Since using a mutex thread can decrease access performance, we decided to execute atomic operations on memory containing the thread state information. We use a mutex thread to synchronize the enqueueing of tasks into the priority queue.

@@

Since $openTaskCount$ has to be altered from different threads the required decrements are ensured to be atomic. This is in opposition to the usual goal of slot-maps to improve cache locality. Atomics on x86 usually prevent their whole cache line from being used while they are changed. Thus a sparse distribution of slots should reduce collisions. To quantify this difference, the baseline implementation searches for free slots linearly from the beginning, while an alternative approach tries to distribute the slots over multiple cache lines. In Figure \ref{fig10}, we show implementations for both search approaches.

@@

\subsection{Managing Thread States}
As mentioned before, accessing the same memory location from different threads may lead to race conditions, which means the program behavior depends on the accessing order of the threads. For example, while one thread updates data at a specific memory location, another thread writes into the data, which results to inconsistency. In Section \ref{sec:pre}, we presented two approaches to handle multithreaded access to the same memory location, mutex threads and atomic operations and we state where those approaches are used in our implementation. In this subsection, we focus on atomic operations to manage thread states. We consider the following steps for a thread to update its state. At first, the current state has to be checked and, depending on the current state, it has to be updated. If these steps are implemented using an if statement and an update function, the complete procedure is not atomic. Consequently, another thread can update the state between check of the current state and update. To avoid this, we use the function \emph{atomic\_compare\_exchange\_strong} \cite{atomicstrong} of the stdatomic library. This function combines the check and update in one atomic procedure. We implemented every thread state access with this function to ensure atomic thread state updates.

@@

\begin{table}[htbp]
	\caption{Thread statistics with their internal names and descriptions}
	\begin{center}
		\begin{tabular}{ l l }
			\hline
			\textbf{Statistic name}&\textbf{Description}\\
			\hline
			waiting\_time & Time the thread spend waiting (ms) \\
			busy\_time & Time the thread spends executing tasks (ms)\\
			task\_count & Number of tasks the thread has executed \\
			\hline
		\end{tabular}
		\label{tab3}
	\end{center}
\end{table}

@@

For testing purposes we used primarily Linux setups, for one Ubuntu 17.04
(linux 4.13) in a virtual machine environment, second Arch Linux (linux
4.17.3-1) running native, both updated on stable branch. Tests have also been
conducted on Windows 10 and macOS High Sierra machines.

In our setups Sky- and Kabylake as well as Ryzen processors have been used with memory
ranging from 8 to 24 GB of both DDR3-SODIMM and DDR4-SODIMM. 
All benchmarks have been repeated three times and a median value of all
repetitions are being used for evaluation.

@@

\subsection{Validity of Results} 
Before presenting our evaluation results, we
give some preliminaries about the validity of our results. At first, all
measurements are executed on all machines we listed in our experimental
configuration, but in this work we present only the results measured on
Arch Linux with kernel 4.17.3-1 and 24 GB DDR4-SODIMM. We decided to use this
system, since it is running a current kernel version and not in a virtual
environment. Additionally, we use gnuplot to generate diagrams from our results,
consequently, using a windows machine is not optimal.

@@

\subsection{Waiting Time Evaluation}
As mentioned before, instead of implementing explicit scheduling to assign tasks to threads, we use the \emph{first come first serve} approach. Each thread that currently has no task, checks the priority queue to pop a new task in a loop. Since our priority queue manages the memory access from different threads using a mutex thread, the queue is locked while a thread asks for a task. Consequently, in large thread pools, the accessing time of the priority queue may increase along with the waiting time of threads. To examine this behavior with increasing thread pool size, we measure the average waiting time of thread pools with different sizes.

@@

\subsubsection{Benchmark Implementation}
To examine the waiting time behavior of larger thread pools, we use a similar implementation than the \emph{cmp\_pool} method, which creates 10000 tasks at different thread pool sizes. Again, we consider sizes from $4$ to $32$ threads as useful test case and take the average value of three measurements. To measure waiting and completion time, we use the monitoring component of our system and evaluate the thread pool statistics of each instance.

@@

\section{Related Work}\label{sec:related}
Implementing multi threading can be realized using different approaches from spawning new threads to reuse them in thread pools. Pyarali et al. show that a thread pool model can significantly improve system performance and reduce response time \cite{pyarali2001evaluating}. They present different approaches to implement thread pools and show the dependency on use cases of these implementations.

@@

\section{Conclusion and Future Work}\label{sec:future}
In this work, we describe our implementation of a thread pool system. We present in detail the implementation and show our evaluation results containing a comparison against the baseline implementation and the waiting time behavior over different thread pool sizes. Within our implementation, we state the differences of parallel memory access as well as the problem of implementing synchronous waiting for a set of tasks.
